Bayesian Parameter Inference for 1D Nonlinear Stochastic Differential Equation Models

### 2015-08-24 TeXFiles/Manuscript_Albert_Ulzega.tex
draft for internal review

\title[Parameter inference with stochastic models]{Bayesian Parameter Inference for Nonlinear Stochastic Differential Equation Models}

@@

\begin{abstract}
Bayesian statistics has become an indispensable tool in many applied sciences for the purpose of uncertainty analysis.
Inferring parametric uncertainty for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
Here we present a Monte Carlo-based method for Bayesian parameter inference with one dimensional SDE models to tackle the generic problems of calibrating the model to given data time series while simultaneously quantifying the ensuing parametric uncertainty.
%Here, we consider the generic problem of calibrating a one dimensional SDE model to %time series and quantifying the ensuing parametric uncertainty.
We re-interpret the Bayesian posterior distribution for model parameters as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to calculate it.
Depending on the number of discretization points and the number of measurement points the dynamics of this system happen on very different time scales.
Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems. Furthermore the algorithm is highly parallelizable.

@@

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB?
\vspace{2pc}
\noindent{\it Keywords}: Bayesian parameter inference, Hamiltonian Monte Carlo, time series analysis, stochastic differential equations

@@

Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density for the parameters is a {\em path-integral}.
Problems of this kind are commonly solved by means of {\em Monte Carlo} methods that are based on simulating model realizations and comparing them to the data. Algorithms of this kind are {\em particle filters} \cite{chopin_2013_SMC2}. For the case of linear SDEs that are coupled to nonlinear deterministic ordinary differential equations (ODEs), more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
The problem with these simulation-based methods is their inefficiency in the presence of many data points.
One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
Such {\em Approximate Bayes Computations} \cite{Albert_2014_ABC} are relatively easy to apply as they only require us to run the simulator. However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.

Exact inference algorithms of high efficiency can be derived from a reinterpretation of the posterior distribution as the partition function of a statistical mechanical system and by simulating the dynamics of the latter.
Such {\em Hamiltonian Monte Carlo} (HMC) algorithms \cite{duane_1987} incorporate the data points for the suggestion of new parameters, whereby much higher acceptance rates are achieved.
The drawback of these methods is that the model equations need to be known and derivatives have to be calculated. The latter problem, however, is largely remedied by the use of automated differentiation algorithms.

Tuning of HMC algorithms is a non-trivial matter.
Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the numerical integration scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
It has been shown that efficiency is gained when the kinetic term is made dependent on the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
Here we explore a different, computationally simper route.
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happen on very different time scales.
Thus, we employ a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
At least for 1D SDEs we always find a parametrization, which allows to partly uncouple the dominant harmonic part of the Hamiltonian in the form of harmonic oscillators and integrate them analytically.

@@

\section{An Exemplary SDE Model}

The methods we present are generic and applicable to a wide range of SDE models.
However, with later applications in mind and to make things concrete, we present them by means of a simple yet non-trivial model from hydrology.
Consider a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
Furthermore, assume this noise to scale linearly with the system state, $S(t)$, which is the water content in the reservoir.
The model equation is thus given by the SDE
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
Scale-invariance of the model of Eq.~(\ref{sde}) for large $S$ leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed~\cite{thyer_2009_fattails}.

Conceptual models of this kind are commonly used in hydrology, for the purpose of predicting rainfall-runoff behaviour, for natural and urban catchments (see, e.g., \cite{breinholt_2011_SDE}).
Furthermore, by means of the transformation $S(t)=1/n(t)$, eq. (\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors and extensively studied \cite{dutre_1977_SDE, fujisaka_1986_intermittency}.

Here, we consider the input $r(t)$ to be a smooth and nowhere vanishing function. Realistic multi-fractal rain inputs \cite{tessier_1996} will be studied elsewhere.
Indeed, the focus of this work is not on hydrological modeling, but on generic methods for parameter inference with SDE models that are calibrated to observed time-series.
Here we assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors, i.e.,
\begin{equation}\label{data}
  \ln \left( y_s \right)
  =
  \ln \left( \frac{S(t_s)}{K} \right)
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,.
\end{equation}
For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}).
Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters is expressed in terms of a probability distribution.
In general, prior knowledge about parameters, $\bt$, is expressed in terms of a prior probability distribution, $f_{\mbox{\small prior}}(\bt)$, while the posterior distribution, which combines prior knowledge with knowledge gained from data, is calculated by means of Bayes' formula
\begin{equation}\label{Bayes}
  f_{\mbox{\small post}}(\bt|\vc y)
  =
  \frac{f_{\mbox{\small prior}}(\bt)L(\vc y|\bt)}
  {\int f_{\mbox{\small prior}}(\bt')L(\vc y|\bt')d\bt'}\,,
\end{equation}
where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution for model outputs given model parameters, evaluated at the measured data.
Here, we assume prior knowledge to be that parameters must be non-negative, but otherwise we assume the prior to be flat on an area large enough to encompass the area where the likelihood function is significantly different from zero. That is, the posterior is data-driven.

Before we set out to derive the likelihood function from the model equations (\ref{sde}) through (\ref{data}) let us express the parameters and state variable with dimensionless quantities.
The noise parameter $\gamma$ is already dimensionless due to scale-invariance of the noise term.
We replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{\mbox d}{\mbox{dt}}\left[ \ln \left( r(t) \right) \right]
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}
For our algorithm it is important to have the model equation in a form where the noise term does not depend on neither the state variables nor the parameters that have to be inferred. In a one dimensional model this can always be achieved through re-parametrization.

The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, is expressed in the form of a {\em path-integral} as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as the infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as \cite{lau_2007}
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.

We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite action~(\ref{action1}) as,
\begin{eqnarray}\label{action}
{\mathcal S}[{q},\dot q] \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\} \nonumber
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0) \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}  \nonumber
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{eqnarray}

@@

If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior as described above, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as a path-integral as follows
\begin{equation}\label{posterior_pathint}
  f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
On the right-hand side of the above equation, the first term describes the probability distribution of model outputs, given model parameters, rain inputs and a system realization $q(t_s)$. The second term is the probability of the given system realization $q$ and the path-integral extends over all possible system realizations. Instead of undertaking an often prohibitive numerical computation of such integral, we apply a Hamiltonian Monte Carlo method, outlined in the next section, to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.

@@

The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta $p$ and $\pi$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
  \item
  The system is then allowed to evolve in $\left(q,\bt; p,{\pmb\pi}\right)$-phase space for an arbitrary time interval $\tau$ according to a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
  %The system is integrated by means of a volume-preserving and time-reversible %solution of a discretization of Hamilton's equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate. Each new phase space configuration corresponds to a well-defined combination of model parameters $\bt$. Thus, the simulated succession of configurations represent a Markov chain of sampled parameter vectors that naturally leads to the sought probability distribution for the system parameters.

In order to simulate the dynamics of the Hamiltonian (\ref{Hamiltonian}), we first need to discretize the path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
The path-integral (\ref{phaseSpacePathInt}) is then approximated by an ordinary integral, with the approximate path-measure
\begin{equation}
  \mathcal Dp\mathcal Dq
  \approx
  \prod_i dp_i dq_i
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$
\begin{eqnarray}
   K( p,{\pmb\pi}) \approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_q}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt) \approx \frac{\Delta t}{T} \sum_{i=2}^{N}
   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\}  \nonumber
  \\
  +
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  \label{Vdisc}
   \,,
\end{eqnarray}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r(t_{i})/r(t_{i-1}))}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}).

>>>>

\title[Parameter inference with stochastic models]{Bayesian Parameter Inference for 1D Nonlinear Stochastic Differential Equation Models}

@@

\begin{abstract}
Bayesian statistics has become an indispensable tool in many applied sciences, in particular, for the purpose of uncertainty analysis.
Inferring parametric uncertainty, for stochastic differential equation (SDE) models, however, is a computationally hard problem due to the high dimensional integrals that have to be calculated.
%Here we present a Monte Carlo-based method for Bayesian parameter inference with generic one dimensional SDE models to tackle the generic problems of calibrating the model to given data time series while simultaneously quantifying the ensuing parametric uncertainty.
Here, we consider the generic problem of calibrating a one dimensional SDE model to time series and quantifying the ensuing parametric uncertainty.
We re-interpret the Bayesian posterior distribution, for model parameters, as the partition function of a statistical mechanical system and employ a Hamiltonian Monte Carlo algorithm to sample from it.
Depending on the number of discretization points and the number of measurement points the dynamics of this system happens on very different time scales.
Thus, we employ a multiple time scale integration together with a suitable re-parametrization to derive an efficient inference algorithm.
While the algorithm is presented by means of a simple SDE model from hydrology, it is readily applicable to a wide range of inference problems. Furthermore the algorithm is highly parallelizable.

@@

%Uncomment for PACS numbers title message
\pacs{00.00, 20.00, 42.10}
% Keywords required only for MST, PB, PMB, PM, JOA, JOB?
\vspace{2pc}
\noindent{\it Keywords}: Bayesian parameter inference, Hamiltonian Monte Carlo, time series analysis, stochastic differential equations, path-integrals

@@

Bayesian parameter inference for SDE models is computationally very expensive, as the posterior probability density for the parameters is a {\em path-integral}.
Problems of this kind are commonly solved by means of {\em Monte Carlo} methods that are based on simulating model realizations and comparing them to the data. Algorithms of this kind are {\em particle filters} \cite{chopin_2013_SMC2}. For the case of linear SDEs that are coupled to nonlinear deterministic ordinary differential equations (ODEs), more efficient algorithms of this kind can be derived \cite{tomassini_2009_smoothing, reichert_2009_timedepParameters}.
The problem with these simulation-based methods is their inefficiency in the presence of many data points.
One solution is to map the output space to a smaller dimensional space of {\em summary statistics}, and acccept/reject proposed model parameters depending on how compatible associated model runs are with the data in terms of these summary statistics.
Such {\em Approximate Bayes Computations} \cite{marin_2012_ABC, Albert_2014_ABC} are relatively easy to apply as they only require us to run the simulator. However, it is largely an unsolved problem how to choose the summary statistics so as to achieve a sufficient approximation of the posterior parameter distribution.

Exact inference algorithms of high efficiency can be derived from a reinterpretation of the posterior distribution as the partition function of a statistical mechanical system and by simulating the dynamics of the latter.
Upon discretizing time of the original problem we are led to the statistical mechanics of a polymer with harmonic bonds in an exterior potential. The measurements are interpreted as an additional exterior potential acting on the measurement beads of the polymer only.
Furthermore, model parameters are interpreted as additional degrees of freedom coupling to all the beads of the polymer.
To simulate the dynamics of this system we apply the {\em Hamiltonian Monte Carlo} (HMC) algorithm \cite{duane_1987}, which combines Molecular Dynamics (MD) \cite{alder_1959_MD, rahman_1964_MD} with the Metropolis algorithm \cite{metropolis_1953}.
Compared to filtering methods, we achieve much higher acceptance rates since data points are already used for the suggestion of new parameters, and thus model realizations that are incompatible with the data are never even considered.
The drawback of these methods is that the model equations need to be known and derivatives have to be calculated. The latter problem, however, is largely remedied by the use of automated differentiation algorithms.

Tuning of HMC algorithms is a non-trivial matter.
Two sorts of tuning parameters need to be adjusted: (i) those of the {\em kinetic energy} of the statistical mechanical system and (ii) those of the {\em numerical integration} scheme of Hamilton's equation in the molecular dynamics part of the HMC algorithm.
It has been shown that efficiency is gained when the kinetic term is made dependent on the configuration of the statistical mechanical system \cite{girolami_2011_HMC}.
Here we explore a different, computationally simper route.
Depending on the number of discretization points needed to approximate the original SDE system and the number of measurement points, the dynamics of the statistical mechanical system happens on very different time scales.
Thus, we employ a multiple time scale integration technique for the simulation of the statistical mechanical system \cite{tuckerman_1993}.
At least for 1D SDEs we always find a parametrization, which, to some extent, decouples harmonic modes in between measurement points from both the measurement points and the model parameters, and allows for an efficient analytical solution of their dynamics.

@@

\section{An Exemplary Inference Problem}

The methods we present are generic and applicable to a wide range of SDE models.
However, with later applications in mind and to make things concrete, we present them by means of a simple yet non-trivial model from hydrology.
Consider, for simplicity, a hydrological catchment whose dynamics at the observation time scale is well described by a linear reservoir and whose other processes happen at much shorter time scales so that they can be described by white noise.
Furthermore, assume this noise to scale linearly with the system state, $S(t)$, which is the water content in the reservoir.
The model equation is thus given by the SDE
\begin{equation}\label{sde}
\dot{S}(t) = r(t) - \frac{1}{K}\left(1+\frac{\gamma}{2}\right) S(t)
+
\sqrt{\frac{\gamma}{K}} S(t){\eta}(t)\,,
\end{equation}
where $r(t)$ denotes the time varying rain input and $\eta(t)$ denotes white noise, i.e.,
\begin{equation}\label{whitenoise}
\langle\eta(t)\eta(t')\rangle = \delta(t-t')\,.
\end{equation}
Eq.~(\ref{sde}) is to be understood in the Stratonovich sense~\cite{stratonovich_1968}.
Our parametrization is such that, for constant rain input $r(t)=r_0$ and in the long-time limit, the mean of $S(t)$ converges to the equilibrium solution of the unperturbed ($\gamma=0$) system, $S_{\eq}=Kr_0$ (see Eq.~(\ref{equilibrium_mean}) below).
Scale-invariance of eq. (\ref{sde}), for large $S$, leads to power law tails in the system state probability distribution (see Eq.~(\ref{inverse_gamma})), which is in line with the observation that errors in hydrological models are often fat-tailed~\cite{thyer_2009_fattails}.

Conceptual models of this kind are commonly used in hydrology, for the purpose of predicting rainfall-runoff behaviour, for natural and urban catchments (see, e.g., \cite{breinholt_2011_SDE}).
Furthermore, by means of the transformation $S(t)=1/n(t)$, eq. (\ref{sde}) turns into a model that has been suggested as a phenomenological description of the dynamics of the neutron density in nuclear reactors \cite{dutre_1977_SDE} and extensively studied, from a theoretical point of view \cite{fujisaka_1986_intermittency}.

Here, we consider the input $r(t)$ to be a smooth and nowhere vanishing function. Realistic multi-fractal rain inputs \cite{tessier_1996} will be studied elsewhere.
Indeed, the focus of this work is not on hydrological modeling, but on generic methods for parameter inference with nonlinear SDE models that are calibrated to observed time-series.
Here we assume the observed time-series, $y_s$, to be the outflow of the reservoir, $S(t)/K$, observed at times $0=t_1<t_2<\dots < t_{n+1}=T$, with multiplicative independent log-normal errors, i.e.,
\begin{equation}\label{data}
  \ln \left( y_s \right)
  =
  \ln \left( \frac{S(t_s)}{K} \right)
  +
  \sigma\epsilon_s\,,\quad s=1,\dots,n+1\,,
\end{equation}
where the $\epsilon_i$ are uncorrelated standard normal errors.

For simplicity, we assume $\sigma$ as well as the input $r(t)$ to be known so that we are left with the task of inferring parameter combinations $(K,\gamma)$ that are compatible with the data given by Eq.~(\ref{data}).
Here, "compatible" is meant in the Bayesian sense, in which knowledge about parameters, $\bt$, is expressed in terms of probability distributions.
Assume we have prior knowledge in the form of a probability distribution, $f_{\mbox{\small prior}}(\bt)$, and measured data, $\vc y$, which is believed to be a realization of the model.
The posterior knowledge, combining prior knowledge with the one acquired from data, is calculated by means of eq.
\begin{equation}\label{Bayes}
  f_{\mbox{\small post}}(\bt|\vc y)
  =
  \frac{f_{\mbox{\small prior}}(\bt)L(\vc y|\bt)}
  {\int f_{\mbox{\small prior}}(\bt')L(\vc y|\bt')d\bt'}\,,
\end{equation}
where $L(\vc y|\bt)$ is the so-called {\em likelihood function}, that is, the probability distribution, for model outputs given model parameters, evaluated at the measured data.
Here, we assume prior knowledge to be that parameters must be non-negative, but otherwise we assume the prior influence to be negligible. That is, we assume the posterior to be data-driven.

Before we set out to derive the likelihood function, from the model equations (\ref{sde}) through (\ref{data}), let us express the parameters and state variable with dimensionless quantities.
The noise parameter $\gamma$ is already dimensionless due to scale-invariance of the noise term.
We replace the state variable $S(t)$ and parameter $K$ by dimensionless quantities $q(t)$ and $\beta$, respectively, by means of the transformations
\begin{equation}
  \beta=\sqrt{\frac{T\gamma}{K}}\,,\quad
  S(t)=\frac{T\gamma r(t)}{\beta^2}e^{\beta q(t)}\,.
\end{equation}
W.r.t. these new  variables and parameters, the model equation (\ref{sde}) becomes
\begin{equation}
  \dot q(t)
  =
  \frac{\beta}{T\gamma}e^{-\beta q(t)}
  -
  \frac{1}{T}\rho(t)
  +
  \frac{1}{\sqrt{T}}\eta(t)\,,
\end{equation}
with
\begin{equation}
  \rho(t)
  =
  \frac{T}{\beta}\frac{\mbox d}{\mbox{dt}}\left[ \ln \left( r(t) \right) \right]
  +
  \frac{(2+\gamma)\beta}{2\gamma}\,.
\end{equation}
For our algorithm it is important to have the model equation in a form where the noise term does neither depend on the state variables nor the parameters that have to be inferred. In a one dimensional model this can always be achieved through re-parametrization.

The probability $P(q_1,T|q_0,0)$ of finding the system in a state $q_1$ at time $t = T$ given that it was in an initial state $q_0$ at time $t = 0$, is expressed in the form of a {\em path-integral} as
\begin{equation}\label{pathint}
P(q_1,T|q_0,0)
=
\frac{1}{Z}
\int
e^{-{\mathcal S}[q,\dot q]}
\delta(q(T)-q_1)
\delta(q(0)-q_0)
\mathcal{D}q \,,
\end{equation}
where the integral extends over all paths $q:[0,T]\rightarrow \mathbb R$.
The path-measure $\mathcal Dq$ is formally written as the infinite product
\begin{equation}\label{pathmeasure_q}
{\mathcal Dq}
=
\prod_{t}
dq(t)\,.
\end{equation}
The {\em action} is a functional on the space of paths and reads as \cite{lau_2007}
\begin{equation}\label{action1}
{\mathcal S}[{q},\dot q]
=
\frac{1}{T}
\int_0^T \dt \left\{
\frac{1}{2}
\left(
    T\dot q(t)
    +
    \rho(t)
    -
    \frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2
    -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
\right\} \,.
\end{equation}
Note that the action includes the Jacobian that arises when changing coordinates from ${\eta(t)}$ to $q(t)$.
We introduce the time-dependent {\em Hamiltonian}
\begin{equation}\label{H}
  \mathcal{H}(q,t)
  =
  \frac{1}{\gamma}e^{-\beta q}+q\rho(t)\,,
\end{equation}
and rewrite action~(\ref{action1}) as,
\begin{eqnarray}\label{action}
{\mathcal S}[{q},\dot q] \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    T\frac{\partial \mathcal{H}(q,t)}{\partial t} -
    \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
     %\frac{(2+\gamma)\beta^2}{4\gamma}
\right\} \nonumber
\\
+ \mathcal{H}(q(T),T) - \mathcal{H}(q(0),0) \nonumber
\\
= \frac{1}{T}
\int_0^T dt\left\{
    \frac{1}{2}
    T^2\dot q^2(t) +
    \frac{1}{2}
    \left(\rho(t)-\frac{\beta}{\gamma}e^{-\beta q(t)}\right)^2 -
    Tq(t)\dot\rho(t) -
     \frac{\beta^2}{2\gamma}e^{-\beta q(t)}
%    \frac{(2+\gamma)\beta^2}{4\gamma}
\right\}  \nonumber
\\
+
    \frac{1}{\gamma}e^{-\beta q(T)}+q(T)\rho(T)
   -\frac{1}{\gamma}e^{-\beta q(0)}-q(0)\rho(0)
\,.
\end{eqnarray}

@@

If we denote the parameter vector $\bt=(\beta,\gamma)^T$ and assume a flat prior, the posterior (\ref{Bayes}), as a function of $\bt$, is proportional to the likelihood function, which is expressed as the path-integral
\begin{equation}\label{posterior_pathint}
  f_{\mbox{\small post}}(\bt | \vc y)
  \propto
  \int
  \exp\bigg[
    -\frac{1}{2}
    \sum_{s=1}^{n+1}
    \frac
    {(\ln(y_s/r(t_s))-{\beta q(t_s)})^2}
    {\sigma^2} -{{\mathcal S}}[{q},\dot q]
    %-\ln(K\gamma)
  \bigg]
  {\mathcal Dq}
  \,.
\end{equation}
The first term in the exponent describes the log-probability distribution of model outputs, given model parameters, inputs and a system realization $q(t_s)$. The second term is the log-probability of a system realization $q(t)$ and the path-integral extends over all possible system realizations. Instead of undertaking an often prohibitive numerical computation of such integral, we apply a Hamiltonian Monte Carlo method, outlined in the next section, to sample parameter vectors from a joint distribution of system realizations and model parameters given by an appropriate discretization of the action of the path-integral.

@@

The HMC method, which is a combination of the {\em Metropolis algorithm} \cite{metropolis_1953} and {\em molecular dynamics} methods \cite{alder_1959_MD, rahman_1964_MD}, iterates the following steps:
\begin{enumerate}
  \item
  Momenta $p(t)$ and $\pi$ are sampled from the Gaussian distributions defined by Eq.~(\ref{K}).
  \item
  The system is then allowed to evolve in $\left(q,\bt; p,{\pmb\pi}\right)$-phase space for an arbitrary time interval $\tau$ according to a volume-preserving and time-reversible solution of a discretization of Hamilton's equations.
  %The system is integrated by means of a volume-preserving and time-reversible %solution of a discretization of Hamilton's equations.
  \item
  The discretization error on the energy preservation due to the previous step is corrected by a Metropolis acceptance/rejection step.
\end{enumerate}
The last step is the standard Metropolis algorithm, while the first two steps allow us to make arbitrarily large jumps in phase space while maintaining an arbitrarily large acceptance rate. Each new phase space configuration is associated with a combination of model parameters $\bt$, which is compatible with the data in the Bayesian sense. Thus, the parameter marginal of the simulated Markov chain of configurations represents a sample of the posterior probability distribution.

In order to simulate the dynamics of Hamiltonian (\ref{Hamiltonian}), we first need to discretize path-integral (\ref{phaseSpacePathInt}).
Therefore, let us assume that the measurement time points $\left\{ y_s \right\}_{s=1,\dots, n+1}$ of the time series (\ref{data}) are equidistantly distributed on the time interval $[0,T]$, with $t_1=0$ and $t_{n+1}=T$.
Each interval between two consecutive data points is further partitioned into $j$ bins, such that we have a total of $nj+1=N>>1$ discretization points.
The path-integral (\ref{phaseSpacePathInt}) is then approximated by an ordinary integral, with the approximate path-measure
\begin{equation}
  \mathcal Dp\mathcal Dq
  \approx
  \prod_i dp_i dq_i
\end{equation}
and the discretized versions of $K( p,{\pmb\pi})$ and $V( q,\bt)$ are given by eqs.
\begin{eqnarray}
   K( p,{\pmb\pi}) \approx
   \sum_{i=1}^N
   \frac{ p_i^2}{2m_i}\Delta t
   +
   \sum_{\alpha=1}^2\frac{\pi_\alpha^2}{2m_\alpha}\,,\label{Kdisc}
   \\
   V(q,\bt) \approx \frac{\Delta t}{T} \sum_{i=2}^{N}
   \left\{ \frac{1}{2} T^2 \dot q_i^2 + \frac{1}{2}
     \left( \rho_i-\frac{\beta}{\gamma}e^{-\beta q_i} \right)^2 -
    \frac{\beta^2}{2\gamma} e^{-\beta q_i} - T q_i\dot\rho_i \right\}  \nonumber
  \\
  +
  \frac{1}{\gamma}
  e^{-\beta q_N}
  +
  q_N \rho_{N}
  -
  \frac{1}{\gamma}
  e^{-\beta q_1}
  -
  q_1 \rho_{1}
  \nonumber
  \\
  +
  \sum_{s=1}^{n+1}
  \frac{(\ln(y_s/r_{(s-1)j+1}) - {\beta q_{(s-1)j+1}})^2}{2\sigma^2}
  \label{Vdisc}
   \,,
\end{eqnarray}
with
\begin{equation}
  \dot q_i = \frac{q_i-q_{i-1}}{\Delta t}\,,
  %\quad
%  \bar q_i = \frac{q_i+q_{i-1}}{2}\,,
\end{equation}
and
\begin{equation}\label{rhodisc}
\rho_i = \frac{T}{\beta} \frac{\ln(r(t_{i})/r(t_{i-1}))}{\Delta t}
+
\frac{(2+\gamma)\beta}{2\gamma}
\,,\quad
\dot\rho_i = \frac{\rho_i-\rho_{i-1}}{\Delta t}\,.
\end{equation}
Note that we've neglected terms of order $\mathcal O(N^{-1/2})$ in the action (\ref{Vdisc}).
